---
title: 'Changing names in the tidyverse: An example for many regressions'
author: Abhijit Dasgupta
date: '2017-03-09'
categories:
  - R
slug: changing-names-in-the-tidyverse-an-example-for-many-regressions
---

A collaborator posed an interesting R question to me today. She wanted to do
several regressions using different outcomes, with models being computed on
different strata defined by a combination of experimental design variables. She then just wanted to extract the p-values for the slopes for each of the models, and then
filter the strata based on p-value levels.

This seems straighforward, right? Let's set up a toy example:


```{r, output='asis', message=FALSE}
library(tidyverse)

dat <- as_tibble(expand.grid(letters[1:4], 1:5))
d <- vector('list', nrow(dat))
set.seed(102)
for(i in 1:nrow(dat)){
  x <- rnorm(100)
  d[[i]] <- tibble(x = x, y1 = 3 - 2*x + rnorm(100), y2 = -4+5*x+rnorm(100))
}
dat <- as_tibble(bind_cols(dat, tibble(dat=d))) %>% unnest()
knitr::kable(head(dat), digits=3, format='html')
```



Now we're going to perform two regressions, one using `y1` and one using `y2` as the dependent variables, for each stratum defined by `Var1` and `Var2`.

```{r}
out <- dat %>%
  nest(-Var1, -Var2) %>%
  mutate(model1 = map(data, ~lm(y1~x, data=.)),
         model2 = map(data, ~lm(y2~x, data=.)))
```

Now conceptually, all we do is tidy up the output for the models using the `broom` package, filter on the rows containg the slope information, and extract the p-values, right? Not quite....

```{r}
library(broom)
out_problem <- out %>% mutate(output1 = map(model1, ~tidy(.)),
                              output2 = map(model2, ~tidy(.))) %>%
  select(-data, -model1, -model2) %>%
  unnest()
names(out_problem)
```


We've got two sets of output, but with the same column names!!! This is a problem! An easy solution would be to preface the column names with the name of the response variable. I struggled with this today until I discovered the _secret function_.

```{r}
out_nice <- out %>% mutate(output1 = map(model1, ~tidy(.)),
                           output2 = map(model2, ~tidy(.)),
                           output1 = map(output1, ~setNames(., paste('y1', names(.), sep='_'))),
                           output2 = map(output2, ~setNames(., paste('y2', names(.), sep='_')))) %>%
  select(-data, -model1, -model2) %>%
  unnest()
```

This is a compact representation of the results of both regressions by strata, and we can extract the information we would like very easily. For example, to extract the stratum-specific slope estimates:

```{r}
out_nice %>% filter(y1_term=='x') %>%
  select(Var1, Var2, ends_with('estimate')) %>%
  knitr::kable(digits=3, format='html')
```

