---
title: Selecting columns based on type
author: Abhijit
date: '2019-11-01'
slug: selecting-columns-based-on-type
categories:
  - R
  - Data Science
tags:
  - Data Science
header:
  caption: ''
  image: ''
  preview: yes
---

The `tidyverse` and, in particular, `dplyr`, provides functions to select columns
from a data frame. There are three *scoped functions* available: `select_all`, `select_if` and `select_at`. In this post, we'll look at a particular application of
`select_if`, i.e., capturing the names of numeric variables.

A quick search using Google finds a few solutions to this problem. As an example data set, I'll use the `diamonds` data set from the `ggplot2` package.

```{r setup, include=F}
library(tidyverse)
```


```{r}
names(diamonds)[sapply(diamonds, is.numeric)]
```
 
or, equivalently

```{r}
names(diamonds)[map_lgl(diamonds, is.numeric)]
```

However, there is an elegant (to me) pipeline based solution using `select_if`.

```{r}
diamonds %>% select_if(is.numeric) %>% names()
```

However, the elegance is at the expense of some efficiency.

```{r}
library(microbenchmark)
microbenchmark(names(diamonds)[sapply(diamonds, is.numeric)],
               names(diamonds)[map_lgl(diamonds, is.numeric)],
               diamonds %>% select_if(is.numeric) %>% names()) %>% 
  autoplot()

```

The piped version is almost 50 times slower. 

