---
title: Creating missing values in factors
author: Abhijit
date: '2020-11-04'
categories:
  - R
tags:
  - data munging
slug: creating-missing-values-in-factors
header:
  caption: ''
  image: ''
  preview: yes
---

## Background

I was looking at some breast cancer data recently, and was analyzing
the ER (estrogen receptor) status variable. It turned out that there were three possible outcomes in the data: Positive, Negative and Indeterminate. I had imported this data as a factor, and wanted to convert the Indeterminate level to a missing value, i.e. NA.

My usual method for numeric variables created a rather singular result:

```{r}
x <- as.factor(c('Positive','Negative','Indeterminate'))
x1 <- ifelse(x=='Indeterminate', NA, x)
str(x1)
```

This process converted it to an integer!! Not the end of the world, but not ideal by any means. 

Further investigation revealed two other `tidyverse` strategies.

### `dplyr::na_if`

This method changes the values to NA, but keeps the original level in the factor's levels

```{r}
x2 <- dplyr::na_if(x, 'Indeterminate')
str(x2)
x2
```

### `dplyr::recode`

This method drops the level that I'm deeming to be missing from the factor

```{r}
x3 <- dplyr::recode(x, Indeterminate = NA_character_)
str(x3)
x3
```

This method can also work more generally to change all values not listed to missing values.

```{r}
x4 <- dplyr::recode(x, Positive='Positive', Negative='Negative', 
                    .default=NA_character_)
x4
```

Other strategies are welcome in the comments.
